---
title: "The Rega User Guide"
bibliography: Rega-refs.bib
author: 
  - name: "Robert Ivánek"
    affiliation: 
    - Department of Biomedicine, University of Basel, Basel, Switzerland
    - Swiss Institute of Bioinformatics, Basel, Switzerland
    email: robert.ivanek@unibas.ch
package: "`r BiocStyle::pkg_ver('Rega')`"
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: 
    default
abstract: "Rega provides interface to European Genome-Phenome Archive (EGA)."
vignette: >
  %\VignetteIndexEntry{The Rega User Guide}
  %\VignetteKeywords{API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: 
    console
---

__Last edited:__ 19 January 2024

```{r init, echo=FALSE, results='hide'}
library(knitr)

## 1. issue with too large vignette, maybe optimizing png?
## unfortunately it requires additional system libs -> not feasible
##           `pngquant=""`, `optipng = "-o7"` both in `opts_chunk`
## SOLUTION: might be to use `fig.retina = 1` and only set it to 2 (default)
##           for figures which requires higher resolution
##           or even decrease res `dpi = 72`, also both in `opts_chunk`
## 
## knit_hooks$set(optipng = hook_optipng,
##                pngquant = hook_pngquant)
##
## 2. crop=NULL or FALSE =>  fix vignette rendering based on yihui/knitr#1796
## added also error=FALSE to include_graphics

opts_chunk$set(
  collapse = TRUE,
  tidy = FALSE,
  fig.retina = 1,
  comment = "#>",
  error = FALSE,
  warning = FALSE,
  message = FALSE,
  crop = NULL                           
)

## check the output type
out_type <- opts_knit$get("rmarkdown.pandoc.to")
if (is.null(out_type))
  out_type <- "html"

## add styling
if (out_type == "html") {
  BiocStyle::markdown()
} else if (out_type == "latex") {
  BiocStyle::latex()
}
```

# Introduction

```{r}
#| label: setup
library(tidyverse)
library(Rega)

# set permanent KEY ? 
# 
# `httr2` is built around the notion that the key should live in an environment variable. 
# So the first step is to make your package key available on your local development 
# machine by adding a line to your your user-level `.Renviron` 
# (which you can easily open with `usethis::edit_r_environ())`:

Sys.setenv("REGA_EGA_USERNAME" = "ega-box-2269")
Sys.setenv("REGA_EGA_PASSWORD" = "4f8T_aX-aLP-5obJWEEIShckVjjovgm3")
```


# Get Available Information From EGA

## User

```{r}
#| label: simple_query_user

ega_users()
```

## Enums

To know which values are accepted for the types i.e `status_types`,`study_types`, `biological_sex`..., you can check all available specification listed in the `enums` that are available by using the following command:

```{r}
#| label: simple_query_enum

ega_enums()
```

### Study types

To check for a specific `enum`, for example `study_types`, you can access just access it to see it's possible values:

```{r}
#| label: simple_query_enum_study_types
ega_enums("study_types")
```

### Run File Types

```{r}
ega_enums("run_file_types") 
```

### Accepted Genomes


```{r}
ega_enums("genomes") 
```


## List Submission

```{r}
#| eval: false

ega_submissions()
ega_submissions(submission = "EGA00002556014")
ega_submissions(submission = "1629667")
```

## List Studies

```{r}
#| eval: false
ega_studies()
ega_studies(study="6743")
ega_studies(study="EGAS00001006794")
ega_studies(submission="1299997")
ega_studies(submission="EGA00002309665")
```

## List Samples

```{r}
#| eval: false

ega_samples()
ega_samples(sample="38812")
ega_samples(submission="7290")
ega_samples(submission="EGA00002556509")
```

## List Experiments

```{r}
#| eval: false

ega_experiments()
ega_experiments(experiment="6971")
ega_experiments(submission="7290")
ega_experiments(submission="EGA00002556509")
```


## List Runs

```{r}
#| eval: false

ega_runs()
ega_runs(run="42191")
ega_runs(submission="7290")
ega_runs(submission="EGA00002556509")
```


## List Analyses

```{r}
#| eval: false

ega_analyses()
ega_analyses(submission="7290")
ega_analyses(submission="EGA00002556509")
```


## List Datasets

```{r}
#| eval: false

ega_datasets()
ega_datasets(submission="1300293")
ega_datasets(submission="EGA00002309961")
```


# Create A Complete New Submission

- submission
- study
- samples
- experiments
- runs
- analyses
- datasets

## Create Submission

```{r}
#| eval: false

new_submission <- ega_create_submission(title="Test Submission API", 
                                        description="Testing API from R")
new_submission
```

### Update Submission

```{r}
#| eval: false

new_submission <- ega_update_submission(submission=new_submission$provisional_id,
                                        title="Test Submission API (update)", 
                                        description="Testing API from R (update)")
new_submission
```

### Rollback Submission

```{r}
#| eval: false

new_submission <- ega_rollback_submission(submission=new_submission$provisional_id)
new_submission
```

### Delete Submission

```{r}
#| eval: false

new_submission <- ega_delete_submission(submission=new_submission$provisional_id)
new_submission
```


## Create Study

```{r}
#| eval: false

new_study <- ega_create_study(submission=new_submission$provisional_id,
                              title="Test Study", 
                              description="New Study, Testing API from R",
                              study_type="RNASeq")
new_study
```


### Update Study

```{r}
#| eval: false

new_study <- ega_update_study(study=new_study$provisional_id,
                                   title="Test Study (updated)", 
                                   description="New Study, Testing API from R (updated)",
                                   study_type="RNASeq")
new_study
```


### Rollback Study

```{r}
#| eval:  false

new_study <- ega_rollback_study(study=new_study$provisional_id)
new_study
```


### Delete Study

```{r}
#| eval:  false

new_study <- ega_delete_study(study=new_study$provisional_id)
new_study
```



## Create Sample

```{r}
#| eval: false
new_sample <- ega_create_sample(submission=new_submission$provisional_id,
                                alias="My unique sample alias 1",
                                biological_sex="male",
                                phenotype="nose",
                                subject_id="192873366738836788")

new_sample
```

### Update Sample

```{r}
#| eval: false

new_sample <- ega_update_sample(sample=new_sample$provisional_id,
                                alias="My unique sample alias 1 (upd)",
                                biological_sex="male",
                                phenotype="nose",
                                subject_id="192873366738836788")
new_sample
```


### Rollback sample

```{r}
#| eval:  false

new_sample <- ega_rollback_sample(sample=new_sample$provisional_id)
new_sample
```


### Delete sample

```{r}
#| eval:  false

new_sample <- ega_delete_sample(sample=sample$provisional_id)
new_sample
```


## Create Experiment

```{r}
#| eval: false
new_experiment <-  ega_create_experiment(submission=new_submission$provisional_id,
                                         design_description = "My experiment design",
                                         study_provisional_id = new_study$provisional_id,
                                         instrument_model_id = 1,
                                         library_layout ="SINGLE",
                                         library_strategy = "WGS",
                                         library_source = "GENOMIC",
                                         library_selection ="RANDOM")

new_experiment
```

### Update experiment

```{r}
#| eval: false

new_experiment <- ega_update_experiment(experiment=new_experiment$provisional_id,
                                        design_description = "My experiment design (updated)",
                                        study_provisional_id = new_study$provisional_id,
                                        instrument_model_id = 1,
                                        library_layout ="SINGLE",
                                        library_strategy = "WGS",
                                        library_source = "GENOMIC",
                                        library_selection ="RANDOM")
new_experiment
```


### Rollback experiment

```{r}
#| eval:  false

new_experiment <- ega_rollback_experiment(experiment=new_experiment$provisional_id)
new_experiment
```


### Delete experiment

```{r}
#| eval:  false

new_experiment <- ega_delete_experiment(experiment=new_experiment$provisional_id)
new_experiment
```

## Create Run

```{r}
#| eval: false

new_run <- ega_create_run(submission=new_submission$provisional_id, 
                          run_file_type="fastq", 
                          files=list("75243", "75406"), 
                          experiment_provisional_id=new_experiment$provisional_id, 
                          sample_provisional_id=new_sample$provisional_id)
new_run
```



```{r}
#| eval: false

new_run <- ega_update_run(run=new_run$provisional_id, 
                          run_file_type="fastq", 
                          files=list("75243", "75406"), 
                          experiment_provisional_id=new_experiment$provisional_id, 
                          sample_provisional_id=new_sample$provisional_id)
new_run
```


### Rollback run

```{r}
#| eval:  false

new_run <- ega_rollback_run(run=new_run$provisional_id)
new_run
```


### Delete run

```{r}
#| eval:  false

new_run <- ega_delete_run(run=new_run$provisional_id)
new_run
```

## Create Analysis


```{r}
#| eval: false

new_analysis <- ega_create_analysis(submission=new_submission$provisional_id, 
                                    title="My first analysis",
                                    description = "Description of my first analysis",
                                    analysis_type="REFERENCE ALIGNMENT", 
                                    files=list("188544"),
                                    study_provisional_id = new_study$provisional_id,
                                    experiment_provisional_ids = list(new_experiment$provisional_id),
                                    sample_provisional_ids = list(new_sample$provisional_id),
                                    experiment_types = list("Whole genome sequencing"),
                                    genome_id = 27)
new_analysis
```

### Update Analysis

```{r}
#| eval: false

new_analysis <- ega_update_analysis(analysis=new_analysis$provisional_id, 
                          title="My first analysis (update)",
                                    description = "Description of my first analysis",
                                    analysis_type="REFERENCE ALIGNMENT", 
                                    files=list("188544"),
                                    study_provisional_id = new_study$provisional_id,
                                    experiment_provisional_ids = list(new_experiment$provisional_id),
                                    sample_provisional_ids = list(new_sample$provisional_id),
                                    experiment_types = list("Whole genome sequencing"),
                                    genome_id = 27)
new_analysis
```


### Rollback analysis

```{r}
#| eval:  false

new_analysis <- ega_rollback_analysis(analysis=new_analysis$provisional_id)
new_analysis
```


### Delete analysis

```{r}
#| eval:  false

new_analysis <- ega_delete_analysis(analysis=new_analysis$provisional_id)
new_analysis
```


## Create dataset

In this case we link with the policy with accession `EGAP00001003010`.

```{r}
#| eval: false

new_dataset <- ega_create_dataset(submission = new_submission$provisional_id,
                                  title = "My dataset title",
                                  description = "My dataset description",
                                  dataset_types = list("Transcriptome profiling by high-throughput sequencing"),
                                  policy_accession_id="EGAP00001003010",
                                  run_provisional_ids=list(new_run$provisional_id),
                                  analysis_provisional_ids = list(new_analysis$provisional_id))

new_dataset
```


### Update Dataset

```{r}
#| eval: false

new_dataset <- ega_update_dataset(dataset = new_dataset$provisional_id,
                                  title = "My dataset title (update)",
                                  description = "My dataset description",
                                  dataset_types = list("Transcriptome profiling by high-throughput sequencing"),
                                  policy_accession_id="EGAP00001003010",
                                  run_provisional_ids=list(new_run$provisional_id),
                                  analysis_provisional_ids = list(new_analysis$provisional_id))
new_dataset
```


### Rollback Dataset

```{r}
#| eval:  false

new_dataset <- ega_rollback_dataset(dataset=new_dataset$provisional_id)
new_dataset
```


### Delete dataset

```{r}
#| eval:  false

new_dataset <- ega_delete_dataset(dataset=new_dataset$provisional_id)
new_dataset
```



# Finalise submission

You can only finalize a submission when the files are in place and all the objects must be used.

```{r}
#| eval: false


# ega_finalise_submission(submission = new_submission$provisional_id,
#                         expected_release_date="2026-01-01")
```

# Submission In One Go

```{r}
#| eval: false

  # check the required columns
  required_cols <- c("alias", "subject_id", "biological_sex", "phenotype")
  if (sum(required_cols %in% colnames(smpl_table)) < 4) {
    stop(paste0("`smpl_table` must contain columns: ", paste(required_cols, collapse=", "), "!"))
  }
  # check that the aliases are unique
  if (nrow(smpl_table) > n_distinct(smpl_table$alias)) {
    stop("`alias` column cannot contain duplicated values!")
  }


# send samples to EGA
smpl_table <- smpl_table |>
  rowwise() |>
  mutate(submission_id = submission_id) |>
  mutate(out = pick(everything())) |>
  mutate(out = list(as.list(out))) |>
  mutate(out = do.call(ega_create_sample, out))

smpl_table$out
```


# Notes Below

```{r}
#| eval: false
#| echo: false

# check the authentication flow

# # token
# token <- oauth_flow_password(
#   client = ega_client,
#   username = ega_username,
#   password = ega_password
# )
# token$access_token

# # cached token
# token <- oauth_token_cached(
#   client = ega_client,
#   flow = oauth_flow_password,
#   flow_params = list(
#     username = ega_username,
#     password = ega_password
#   ),
#   cache_disk = TRUE
# )
# token$access_token
```

