---
title: "The Rega User Guide"
bibliography: Rega-refs.bib
date: "Last edited: `r format(Sys.time(), '%d %B, %Y')`"
author: 
  - name: "Robert Ivánek"
    affiliation: 
    - Department of Biomedicine, University of Basel, Basel, Switzerland
    - Swiss Institute of Bioinformatics, Basel, Switzerland
    email: robert.ivanek@unibas.ch
  - name: "Igor Cervenka"
    affiliation: 
    - Department of Biomedicine, University of Basel, Basel, Switzerland
    - Swiss Institute of Bioinformatics, Basel, Switzerland
    email: igor.cervenka@unibas.ch
package: "`r BiocStyle::pkg_ver('Rega')`"
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: 
    default
abstract: "Rega provides interface to European Genome-Phenome Archive (EGA)."
vignette: >
  %\VignetteIndexEntry{The Rega User Guide}
  %\VignetteKeywords{API}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r options, include=FALSE}
library(knitr)
opts_chunk$set(
    echo = TRUE,
    eval = TRUE,
    message = FALSE,
    warning = FALSE,
    collapse = FALSE,
    comment = NA,
    prompt = FALSE,
    tidy = FALSE,
    dev = "png",
    out.width = "100%"
)

## check the output type
out_type <- opts_knit$get("rmarkdown.pandoc.to")
if (is.null(out_type)) {
    out_type <- "html"
}

## add styling
if (out_type == "html") {
    BiocStyle::markdown()
} else if (out_type == "latex") {
    BiocStyle::latex()
}
```

# Install/Load packages

Install (if necessary) and load all necessary packages

```{r load-package}
# Install CRAN packages
install.packages("devtools")   
library(devtools)
install.packages("httr2") 
library(httr2)

# Install GitHub packages
devtools::install_github("ivanek/Rega")  
library(Rega)


```

# Fill in the submission template

Download the empty MS Excel template from
`inst/extdata/ega_full_template_v2.xlsx` and fill it in according to the
instructions in the 'Instructions' tab.

# Setup credentials

`httr2` is designed around the idea that authentication credentials,
such as API keys or passwords, should be stored in an environment
variable rather than written directly in the code. To do so, you need
first to create a secret package key:

## Create a secret package key

First, create a secret key that will be used to encrypt and decrypt your
password.

```{r generate_key}
# Generate a secret key
httr2::secret_make_key() # returns a string eg. "5h4Yk...X2a"
```

To make this package key available on your local development machine,
store it as the "REGA_KEY" environmental variable it your user-level
`.Renviron` file, which is automatically loaded when R starts.

```{r store_key, eval=FALSE}
usethis::edit_r_environ()
# Add the following line to the .Renviron file
# (replace "5h4Yk...X2a" with your actual key)
REGA_KEY=5h4Yk...X2a
```

Restart R after saving the file so the variable becomes available.

## Encrypt and store your EGA password

Environment variables should not store passwords in plain text. Instead,
use the secret key stored in `REGA_KEY` to encrypt your EGA password.

```{r encrypt_password}
encrypted_password <- httr2::secret_encrypt(
  "<your-ega-password>",
  "REGA_KEY"
) 
encrypted_password # this encrypted string is safe to store.
```

Store the encrypted password as an environment variable in your
`.Renviron` file:

Now you can create an `REGA_EGA_PASSWORD` environmental variable which
will store your EGA password. However, environmental variable aren't
supposed to store passwords in plain text. Therefore, use your secret
key stored as `REGA_KEY` to encrypt your EGA password

This encrypted string is then safe to store as an environment variable
in the `.Renviron` file

```{r store_encrypted_password, eval=FALSE}
usethis::edit_r_environ()
# Add the encrypted password to the .Renviron file
REGA_EGA_PASSWORD=<your-encrypted-password>
```

Restart R after saving the file so the variable becomes available.

# Data submission

## Submission metadata parsing

The default parser parses the bundled `xlsx` template
(`inst/extdata/ega_full_template_v2.xlsx`). Changing the parser
behavior, such as whether to append `c4gh` extension to files or which
sheets to parse from the metadata files is controlled by YAML config
file `inst/extdata/default_parser_params.yaml`. To modify the
parameters, create a local copy of the file and pass it as `param_file`
argument to `default_parser` function.

```{r parse-metadata}
metadata_file <- system.file(
    "extdata/submission_example.xlsx",
    package = "Rega"
)

parsed_metadata <- default_parser(metadata_file)
parsed_metadata
```

## Submission metadata validation

The internal validation is a bit stricter in enforcing uniqueness of
titles and descriptions for individual tables (the API itself doesn't
have this requirement). If the validation only fails in this particular
case, you can still continue with creating the submission. However,
there have been instances of failed reviews due to non-unique titles and
descriptions before.

```{r validate-metadata}
validation_summary <- default_validator(parsed_metadata)
validation_summary
```

## Create API client

Extract EGA API using the bundled YAML specification

```{r extract-api}
api <- extract_api()
```

Create a client using the embedded `httr2` OAuth authentication
(default).

```{r client-with-oauth}
ega <- create_client(api, verbosity = 0)
```

## Run new submission workflow

Specify one of the available workflow function with your parsed metadata
and a client as parameters.

```{r run-workflow, eval=FALSE}
responses <- new_submission(parsed_metadata, ega, "log.yaml")
```

## Other workflows

You can get the entire contents of current submission metadata via
`get_submission` workflow.

```{r get-submission-contents, eval=FALSE}
resp <- get_submission(00001, ega, logfile = "log.yaml")
resp <- get_submission(00001, ega)
```

Or delete the entire contents of current submission metadata via
`delete_submission_contents` workflow or delete the entire submission by
using the `delete_submission` workflow.

```{r delete-submission-contents, eval=FALSE}
resp <- delete_submission_contents(00001, ega)
resp <- delete_submission(00001, ega)
```

**Workflow for updating the submission metadata by `PUT` method is under
development.**

# Authentication

`Rega` package supports authentication through OAuth or Bearer token via
functions `ega_oauth` (default) and `ega_token` respectively.

# Utilities

If you wish to create your own templates for EGA submissions, we provide
a few functions to retrieve properties and enums through API and save
them in text files. We will use the API and the client created above.

Relevant functions include:

-   `get_schemas()`
-   `get_properties()`

# Notes

## Bearer token authentication

For testing, debugging and prototyping purposes, it is possible to
directly use generated bearer token with API when creating the client.
It is then your responsibility of the user to track the validity and
refresh the token as necessary.

```{r client-with-token, eval=FALSE}
bt <- ega_token()
ega <- create_client(api, bt$access_token)

ega$get__enums()
```

# Issues

# Session Info

```{r session-info}
sessionInfo()
```
